---
title: "Exploring data"
format: html
---
## Introduction

This document contains the final analysis for the baggie case paper. The starting point will be the outline of the result section that I have already written, and as analysis reveals insight/new questions, I will add/subtract from there. I will start by calculating some general statistics about the data and the sampling effort. Then, I will look at paired data subset, first by creating a data frame that will be used in all the possible analyses for those data. After, I will continue with the analysis for the whole project. 

## Set up

Let's load in the necessary packages and the merged data file (containing both deployment data and BirdNET output).

```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(pander)
merged <- read_csv(here("data/merged_deployment_birdnet.csv"))
birdnet <- read_csv(here("data/birdnet_results.csv"))
deploy <- read_csv("data/Updated_baggie_case_deployments.csv")
history <- read_csv("data/acoustic_device_deployments.csv")
```

Let's take a look at the structure:

```{r}
glimpse(merged)
```


## Summary Statistics

There are some variables to add to the data frame:

 - record_len: stores the length of actual recording (using the deployment date and last recording date)

 - battery_life: stores the expected battery life based on the config file (same for every AM)

 - *gb_full: amount of time expected to fill the sd card based on card size and config file*
     - *Would have added sd_full variable but the battery seems like the limiting factor (regardless of the 64 GB or 256 GB cards)*

```{r}
# save recording length and battery life as new variables
merged <- merged |> mutate(dropoff_date = dmy(dropoff_date),
last_record = mdy(last_record),
record_len = last_record - dropoff_date,
battery_life = rep(2500/60, nrow(merged)), # 2500 mAh with a daily usage of 60 mAh based on config
gb_per_d = rep(1.162, nrow(merged))) |> # GB used per day based on config
filter(deployment_num != "AM0100") # the known AM with user error

# add these same variables to the deployment data (also change pick up data type):
deploy <- deploy |> mutate(dropoff_date = dmy(dropoff_date),
last_record = mdy(last_record),
record_len = last_record - dropoff_date,
battery_life = rep(2500/60, nrow(deploy)),
gb_per_d = rep(1.162, nrow(deploy)),
pickup_date = dmy(pickup_date))|> 
filter(deployment_num != "AM0100") 

glimpse(merged)
glimpse(deploy)
```


Total nights of deployment and nights of recording:

```{r}
deploy |> summarise(tot_deploy = sum(pickup_date - dropoff_date, na.rm = T),
tot_rec = sum(record_len, na.rm = T)) |> pander()
```

Total ids and total species richness:

```{r}
birdnet |> summarise(tot_ids = n(),
tot_sp = n_distinct(species)) |> pander()
```

Total identifications and species richness per AM:

```{r}
merged |> group_by(deployment_num) |> 
    summarise(num_ids = sum(!is.na(probability)),
    num_sp = n_distinct(species, na.rm = T)) |> 
        pander()
```

#### Baggie vs Case Comparisons

Total identifications and species richness:

```{r}
merged |> group_by(AM_housing) |>
    summarise(total_ids = sum(!is.na(probability)),
    total_sp = n_distinct(species, na.rm = T)) |> 
        pander() 
```

Mean and sd:

```{r}
merged |> group_by(AM_housing, deployment_num) |> 
    summarise(total_ids = sum(!is.na(probability)),
    total_sp = n_distinct(species, na.rm = T)) |> 
        group_by(AM_housing) |> 
        summarise(mean_ids = mean(total_ids),
        sd_ids = sd(total_ids),
        mean_sp = mean(total_sp),
        sd_sp = sd(total_sp)) |> 
            pander()
```

## Paired Analysis

Here the analysis will focus on the subset of deployments that included two AMs at the same site.

Let's create the data frame needed for the paired analysis:

```{r}
# create a dataset that has the species richness and total IDs for each AM
rich_and_ids <- merged |> filter(paired == "y") |> 
    group_by(deployment_num) |> 
    summarise(id_count = sum(!is.na(probability)),
    species_count = n_distinct(species, na.rm = T))

# merge the deployment 
paired <- deploy |> filter(paired == "y") |> 
    select(deployment_num, AM_housing, paired_site) |> 
    left_join(rich_and_ids, join_by("deployment_num" == "deployment_num")) |> 
    mutate(across(c(id_count, species_count), ~replace_na(., 0)))|> 
    pivot_wider(id_cols = paired_site, names_from = AM_housing,
    values_from = c(id_count, species_count)) |> 
        mutate(diff_detect = id_count_Case - id_count_Ziplock,
        diff_SR = species_count_Case - species_count_Ziplock)
```

#### Statistical test

Since we have a low number of replicates, we will use a Wilcoxon Signed Rank Test

```{r}
wilcox.test(paired$id_count_Case, paired$id_count_Ziplock)

wilcox.test(paired$species_count_Case, paired$species_count_Ziplock)
```

Not significant, but there still might be a trend

#### Plot

species richness
```{r}
# data for plot
paired_plot <- deploy |> filter(paired == "y") |> 
    select(deployment_num, AM_housing, paired_site) |> 
    left_join(rich_and_ids, join_by("deployment_num" == "deployment_num")) |> 
    mutate(across(c(id_count, species_count), ~replace_na(., 0)))

# species richness
ggplot(paired_plot, aes(AM_housing, species_count, group = paired_site))+
    geom_point(aes(color = AM_housing))+
    geom_line()+
    theme_minimal()

# total ids
ggplot(paired_plot, aes(AM_housing, id_count, group = paired_site))+
    geom_point(aes(color = AM_housing))+
    geom_line()+
    theme_minimal()
```

## Durability Analysis

Due to the estimation by the configuration app based on our config file, we know that the batteries will last around 41.67 days and we used 1.162 GB on the SD card per day. Since the SD cards were either 64 GB or 256 GB, the batteries are the clear limiting factor on how long we would predict the AM to last. Therefore, for our durability analysis, our response variable will be the percentage of days that the AM lasted based on our expected battery life.

Let's add this variable to the deployment dataset:


```{r}
durability <- deploy |> 
    mutate(record_len = 
    replace(record_len, deployment_num == "AM0057", 0), # AM did not record at all (coded as 0)
    percentage = record_len/battery_life)
```

#### Basic statistics

```{r}
durability |> 
    group_by(AM_housing) |> 
    summarise(mean_percent = mean(percentage),
    sd_percent = sd(percentage)) |> 
        pander()
```

What if we remove the failed AM?

```{r}
durability |>
    filter(deployment_num != "AM0057") |> 
    group_by(AM_housing) |> 
    summarise(mean_percent = mean(percentage),
    sd_percent = sd(percentage)) |>
        pander()
```

The cases lasted for a smaller percentage of the expected recording duration on average!

However, there is less variation in recording duration in the case group (especially after removing the failed AM).

Let's see a visual of these results:

Side-by-side boxplots:

```{r}
ggplot(durability, aes(AM_housing, percentage))+
    geom_boxplot()+
    geom_point()+
    theme_minimal()
```

Very intersting! 

#### Historical Deployments

Perhaps we are looking at the wrong variable. Clearly housing does not seem to affect durability much, but previous deployments might be a much stronger predictor of durability. 

Using the complete deployment history of each audiomoth we can see if the total deployment days influences the 'recording length' of the first and second round of deployments in this experiment.


We need to filter out all the deployments after the baggie/case experiment:
```{r}
last_deploy <- deploy |> 
    filter(dropoff_date == max(dropoff_date)) |> 
    slice(1)

history_cleaned <- history |> 
    select(deployment_num, am_recorder_lab_id_num, dropoff_date, pickup_date, AM_housing) |> 
    mutate(dropoff_date = lubridate::dmy(dropoff_date),
    pickup_date = lubridate::dmy(pickup_date)) |> 
    filter(dropoff_date <= last_deploy$dropoff_date) |> 
    filter(am_recorder_lab_id_num %in% deploy$recorder_lab_id_num)
    
    # only look at AMs that we deployed for the experiment

# look at the all the AM id numbers and the number of deployments and total 
history_cleaned |> group_by(am_recorder_lab_id_num) |> 
    summarise(tot_days = sum(pickup_date - dropoff_date),
    n_deploy = n())
```

Regression with total number of days of deployments and recording length on the last deployment:

```{r}
record_len <- durability |> arrange(desc(dropoff_date)) |> 
    group_by(recorder_lab_id_num) |> 
    summarise(first_rec_len = last(record_len),
    last_rec_len = first(record_len))

history_plot <- history_cleaned |> group_by(am_recorder_lab_id_num) |> 
    summarise(tot_days = sum(pickup_date - dropoff_date),
    n_deploy = n()) |> 
    ungroup() |> 
    full_join(record_len, join_by(am_recorder_lab_id_num == recorder_lab_id_num))

# model
hist_mod <- lm(as.numeric(last_rec_len)~as.numeric(tot_days), data = history_plot)
summary(hist_mod)

# plot total number of days deployed
ggplot(history_plot, aes(tot_days, last_rec_len))+
    geom_point()+
    geom_smooth(se = F)+
    theme_minimal()

# plot number of deployments
ggplot(history_plot, aes(n_deploy, last_rec_len))+
    geom_point()+
    geom_smooth(se = F)+
    theme_minimal()
```


