---
title: "Data Analysis"
author: "Aidan Fauth"
format: 
  docx:
    embed-resources: true
# execute: 
#   eval: false
---
## Introduction

This document contains the final analysis for the baggie/case paper. The starting point will be the outline of the result section that I have already written, and as analysis reveals insight/new questions, I will add/subtract from there. I will start by calculating some general statistics about the data and the sampling effort. Then, I will look at paired data subset, first by creating a data frame that will be used in all the possible analyses for those data. After, I will continue with the analysis for the whole project. 

## Set up

Let's load in the necessary packages and the merged data file (containing both deployment data and BirdNET output).

```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(pander)
library(sf)
library(spmodel)
merged <- read_csv(here("data/merged_deployment_birdnet.csv"))
birdnet <- read_csv(here("data/birdnet_results.csv"))
deploy <- read_csv(here("data/Updated_baggie_case_deployments.csv"))
history <- read_csv(here("data/acoustic_device_deployments.csv"))
```

Let's take a look at the structure:

```{r}
glimpse(merged)
```


## Summary Statistics

There are some variables to add to the data frame:

 - record_len: stores the length of actual recording (using the deployment date and last recording date)

 - battery_life: stores the expected battery life based on the config file (same for every AM)

 - *gb_full: amount of time expected to fill the sd card based on card size and config file*
     - *Would have added sd_full variable but the battery seems like the limiting factor (regardless of the 64 GB or 256 GB cards)*

```{r}
# save recording length and battery life as new variables
merged <- merged |> mutate(dropoff_date = dmy(dropoff_date),
pickup_date = dmy(pickup_date),
last_record = mdy(last_record),
record_len = last_record - dropoff_date,
battery_life = rep(2500/60, nrow(merged)), # 2500 mAh with a daily usage of 60 mAh based on config
gb_per_d = rep(1.162, nrow(merged))) |> # GB used per day based on config
filter(deployment_num != "AM0100" & deployment_num != "AM0057") |> # the known AM with user error
mutate(deploy_len = pickup_date - dropoff_date)

# add these same variables to the deployment data (also change pick up data type):
deploy <- deploy |> mutate(dropoff_date = dmy(dropoff_date),
pickup_date = dmy(pickup_date),
last_record = mdy(last_record),
record_len = last_record - dropoff_date,
battery_life = rep(2500/60, nrow(deploy)),
gb_per_d = rep(1.162, nrow(deploy)))|> 
filter(deployment_num != "AM0100" & deployment_num != "AM0057") |>
mutate(deploy_len = pickup_date - dropoff_date)

glimpse(merged)
glimpse(deploy)
```


No anthroprogenic noise:

```{r}
anthro <- c("Fireworks_Fireworks", "Siren_Siren", "Engine_Engine", "Gun_Gun", 
  "Human vocal_Human vocal", "Human non-vocal_Human non-vocal", "Dog_Dog")

cleaned <- merged |> filter(!(species %in% anthro))
```

Total nights of deployment and nights of recording:

```{r}
deploy |> summarise(tot_deploy = sum(pickup_date - dropoff_date, na.rm = T),
tot_rec = sum(record_len, na.rm = T)) |> pander()
```

Total ids and total species richness:

```{r}
birdnet |> summarise(tot_ids = n(),
tot_sp = n_distinct(species)) |> pander()
```

Total identifications and species richness per AM:

```{r}
merged |> group_by(deployment_num) |> 
    summarise(num_ids = sum(!is.na(probability)),
    num_sp = n_distinct(species, na.rm = T)) |> 
        pander()
```

#### Baggie vs Case Comparisons

Total identifications and species richness:

```{r}
merged |> group_by(AM_housing) |>
    summarise(total_ids = sum(!is.na(probability)),
    total_sp = n_distinct(species, na.rm = T)) |> 
        pander() 
```

Mean and sd:

```{r}
merged |> group_by(AM_housing, deployment_num) |> 
    summarise(total_ids = sum(!is.na(probability)),
    total_sp = n_distinct(species, na.rm = T)) |> 
        group_by(AM_housing) |> 
        summarise(mean_ids = mean(total_ids),
        sd_ids = sd(total_ids),
        mean_sp = mean(total_sp),
        sd_sp = sd(total_sp)) |> 
            pander()
```

#### Species List Comparison

```{r}
Species <- unique(merged$species)

case_species <- merged |> filter(AM_housing == "Case")

zip_species <- merged |> filter(AM_housing == "Ziplock")

Case <- c()
Ziplock <- c()

for (i in Species){
    if (i %in% case_species$species & i %in% zip_species$species){
        Case <- c(Case, "X")
        Ziplock <- c(Ziplock, "X")
    } else if (i %in% case_species$species){
        Case <- c(Case, "X")
        Ziplock <- c(Ziplock, "")
    }else {
        Case <- c(Case, "")
        Ziplock <- c(Ziplock, "X")
    }

}

b_c_list <- tibble(Species, Case, Ziplock)

b_c_list |> arrange(Species)|> 
    separate(Species, into = c("Speices", "Common Name"), sep = "_")|> 
    pander()
```

## Paired Analysis

Here the analysis will focus on the subset of deployments that included two AMs at the same site.

Let's create the data frame needed for the paired analysis:

### IDs

```{r}
# create a dataset that has the species richness and total IDs for each AM
ids <- merged |> 
    group_by(deployment_num, day(date_time)) |> 
    summarise( id_day = sum(!is.na(probability))) |> 
    ungroup() |> 
    group_by(deployment_num) |> 
    summarise(id_mean = mean(id_day)) |> 
    ungroup()

# merge the deployment 
paired_ids <- deploy |> filter(paired == "y") |> 
    select(deployment_num, AM_housing, paired_site, record_len) |> 
    left_join(ids, join_by("deployment_num" == "deployment_num")) |> 
    pivot_wider(id_cols = paired_site, names_from = AM_housing,
    values_from = id_mean) |> 
        mutate(diff_detect = Case - Ziplock)
```

#### Statistical test

Since we have a low number of replicates, we will use a Wilcoxon Signed Rank Test

```{r}
wilcox.test(paired_ids$Case, paired_ids$Ziplock, paired = TRUE)
```

Not significant, but there still might be a trend

#### Plot

species richness
```{r}
# data for plot
paired_plot_ids <- deploy |> filter(paired == "y") |> 
    select(deployment_num, AM_housing, paired_site, record_len) |> 
    left_join(ids, join_by("deployment_num" == "deployment_num"))

# total ids
ggplot(paired_plot_ids, aes(AM_housing, id_mean, group = paired_site))+
    geom_point(aes(color = AM_housing))+
    geom_line()+
    labs(x = "Housing Type",
        y = "Average detections per day",
        color = "Housing Type")+
    scale_colour_viridis_d()+
    theme_minimal()+
    theme(legend.position = "none")
```

### Species Richness

Remove anthroprogenic noises

```{r}
# create a dataset that has the species richness and total IDs for each AM
rich_cleaned <- cleaned |>
    group_by(deployment_num, day(date_time)) |> 
    summarise(sr_day = n_distinct(species, na.rm = T)) |> 
    ungroup() |> 
    group_by(deployment_num) |> 
    summarise(sr_mean = mean(sr_day)) |> 
    ungroup()

# merge the deployment 
paired_sr <- deploy |> filter(paired == "y") |> 
    select(deployment_num, AM_housing, paired_site, record_len) |> 
    left_join(rich_cleaned, join_by("deployment_num" == "deployment_num")) |> 
    pivot_wider(id_cols = paired_site, names_from = AM_housing,
    values_from = sr_mean) |> 
        mutate(diff_SR = Case - Ziplock)
```

#### Test
```{r}
wilcox.test(paired_sr$Case, paired_sr$Ziplock, paired = TRUE)
```

#### Plot

```{r}
paired_plot_sr <- deploy |> filter(paired == "y") |> 
    select(deployment_num, AM_housing, paired_site, record_len) |> 
    left_join(rich_cleaned, join_by("deployment_num" == "deployment_num"))

# species richness
ggplot(paired_plot_sr, aes(AM_housing, sr_mean, group = paired_site))+
    geom_point(aes(color = AM_housing))+
    geom_line()+
    labs(x = "Housing Type",
        y = "Average SR per day",
        color = "Housing Type")+
            scale_colour_viridis_d()+
    theme_minimal()+
    theme(legend.position = "none")
```

## Durability Analysis

Due to the estimation by the configuration app based on our config file, we know that the batteries will last around 41.67 days and we used 1.162 GB on the SD card per day. Since the SD cards were either 64 GB or 256 GB, the batteries are the clear limiting factor on how long we would predict the AM to last. Therefore, for our durability analysis, our response variable will be the percentage of days that the AM lasted based on our expected battery life.

Let's add this variable to the deployment dataset:


```{r}
durability <- deploy |> 
    mutate(percentage = record_len/battery_life)
```

#### Basic statistics

```{r}
durability |> 
    group_by(AM_housing) |> 
    summarise(mean = mean(record_len),
    sd = sd(record_len)) |> 
        pander()
```


The cases lasted for a smaller percentage of the expected recording duration on average!

However, there is less variation in recording duration in the case group (especially after removing the failed AM).

### Test

```{r}
# need to only use round two deployments
durability_test <- durability |> 
    select(deployment_num, AM_housing, deploy_len, record_len, percentage) |> 
    filter(deploy_len > 21)

wilcox.test(as.numeric(record_len) ~ AM_housing, data = durability_test)
```

### Plot

```{r}
ggplot(durability_test, aes(AM_housing, record_len))+
    geom_boxplot()+
    geom_jitter(width = 0.03, height = 0)+
    geom_hline(yintercept = 41.67, linetype = "dashed", color = "red")+
    scale_y_continuous(limits = c(0, NA))+
    labs(x = "Housing Type",
    y = "Recording length (days)")+
    theme_minimal()
```

## All Sites

Now for the big finale... we will look at the the entire data set, both paired and upaired, to see if either housing group is significantly better at detecting more species (species richness) or just number of detections.

We have two candidate models: 

1. Multilevel Model: $SR = \beta_0 + \beta_1case\_type + \epsilon$
 - where the random effect is for paired sites
 - assumption is that all unpaired sites are independent

2. $SR = \beta_0 + \beta_1case\_type + \epsilon + \tau$
 - where the random effect is for paired sites
 - $\tau$ is a spatial dependent term with some covariance structure: 
  * $Cov(\tau_i, \tau_j) = \sigma^2 exp(-d_{ij} / \phi)$
 - no point is considered spatially independent

We will use `spmodel` for both models so that we can compare them with the same AIC/BIC.

#### Multilevel Model for SR

First we need to manipulate the the deployment data to have a unique ID for each paried site and re-join with the birdnet output.

```{r}
deploy_mod_df <- deploy |> mutate(pair_id = if_else(paired == 'y',
paste(pmin(deployment_num, paired_with), 
pmax(deployment_num, paired_with), 
sep = "_"), deployment_num)) |> 
mutate(deploy_len = pickup_date - dropoff_date,
Round = if_else(deploy_len <= 21, 
    true = "Round 1", false = "Round 2"))

sr_mod_df <- left_join(deploy_mod_df, rich_cleaned, 
join_by(deployment_num == deployment_num))
```

Convert this data frame into an sf object using the NAD83 projection (EPSG: 5070):

```{r}
sr_mod_sf <- st_as_sf(sr_mod_df, coords = c("longitude", "latitude"), crs = 4326)

sr_mod_sf <- st_transform(sr_mod_sf, crs = 5070)
```

Now we can use splm() to create the model:

```{r}
mixed_mod <- splm(sr_mean ~ AM_housing + Round, data = sr_mod_sf, 
random = ~ 1 | pair_id, spcov_type = "none")
summary(mixed_mod)
```

significant p-value now that we account for round and average per day!

#### Spatial Model for SR

Now we can try the spatial model:

```{r}
spatial_mod <- splm(sr_mean ~ AM_housing + Round, data = sr_mod_sf, 
spcov_type = "exponential")
summary(spatial_mod)
```

Same deal.

 Comparison:

```{r}
glances(mixed_mod, spatial_mod) |> pander()
```

Looks like the the standard mixed model did better and standardizing for recording length seems to have increased the near-significant p-value for the mixed model. 

```{r}
tidy(mixed_mod) |> pander()
```

#### Mixed model for IDs

```{r}
deploy_mod_df <- deploy |> mutate(pair_id = if_else(paired == 'y',
paste(pmin(deployment_num, paired_with), 
pmax(deployment_num, paired_with), 
sep = "_"), deployment_num)) |> 
mutate(deploy_len = pickup_date - dropoff_date,
Round = if_else(deploy_len <= 21, 
    true = "Round 1", false = "Round 2"))

id_mod_df <- left_join(deploy_mod_df, ids, 
join_by(deployment_num == deployment_num))
```

Convert this data frame into an sf object using the NAD83 projection (EPSG: 5070):

```{r}
id_mod_sf <- st_as_sf(id_mod_df, coords = c("longitude", "latitude"), crs = 4326)

id_mod_sf <- st_transform(id_mod_sf, crs = 5070)
```

Now we can use splm() to create the model:

```{r}
mixed_id_mod <- splm(id_mean ~ AM_housing + Round, data = id_mod_sf, 
random = ~ 1 | pair_id, spcov_type = "none")
summary(mixed_id_mod)
```

```{r}
tidy(mixed_id_mod) |> pander()
```

#### Plots

First lets look at a side-by-side boxplot for all the sites

```{r}
overall_plot <- left_join(deploy, rich_cleaned, join_by(deployment_num == deployment_num)) |> 
    mutate(
    paired = fct_recode(paired, "paired" = "y", "not paired" = "n"))

ggplot(overall_plot, aes(AM_housing, sr_mean))+
    geom_boxplot(color = "black")+
    geom_jitter(width = 0.04, height = 0, aes(color = paired))+
    scale_color_manual(values = c("paired" = "red", "not paired" = "black"))+
    labs(x = "Housing Type",
    y = "Average SR per day",
    color = "Site Type")+
    theme_minimal()

id_plot <- left_join(deploy, ids, join_by(deployment_num == deployment_num)) |> 
    mutate(
    paired = fct_recode(paired, "paired" = "y", "not paired" = "n"))

ggplot(id_plot, aes(AM_housing, id_mean))+
    geom_boxplot(color = "black")+
    geom_jitter(width = 0.04, height = 0, aes(color = paired))+
    scale_color_manual(values = c("paired" = "red", "not paired" = "black"))+
    labs(x = "Housing Type",
    y = "Average species detections per day",
    color = "Site Type")+
    theme_minimal()
```


## Species Accumulation

```{r}
species_accum <- tibble(
  deployment_num = character(),
  date = POSIXct(),
  num_sp_today = integer(),
  num_sp_shared_yesterday = integer(),
  num_new_sp = integer(),
  cum_num_species = integer()
)
```

```{r}
cleaned <- cleaned |> mutate(date = date(date_time))

deployments <- unique(cleaned$deployment_num)
#empty list to hold lists of species per detector
#create list using deployments so that each list element is named to a particular deployment

sp_list_by_deployment <- sapply(deployments, function(x) NULL)

#second empty list to hold lists of cumulative species per detector
cum_sp_list_by_deployment <- sapply(deployments, function(x) NULL)

#start loop
for(i in 1:length(deployments)){
  my_deployment <- deployments[i]

  temp_df <- cleaned|> filter(deployment_num == my_deployment) |> arrange(deployment_num, date_time)
  
  #second loop to isolate days
  days <- unique(temp_df$date)
  #empty list to hold species lists
  sp_list_by_day <- vector(mode = "list", length = length(days))
  
  #second empty list to hold cumulative species lists
  cum_sp_by_day <- vector(mode = "list", length = length(days))
  
  for(a in 1:length(days)){
    my_date <- days[a]
    
    day_df <- temp_df |> filter(date == my_date)
    num_sp_today <- n_distinct(day_df$species)
    
    sp_list_by_day[[a]] <- unique(day_df$species)
    if (a == 1){
      num_sp_shared_yesterday <- 0} else {
      num_sp_shared_yesterday <- n_distinct(intersect(sp_list_by_day[[a-1]], sp_list_by_day[[a]]))
      }
    if (a == 1){
      num_new_sp <- num_sp_today } else {
        num_new_sp <- length(setdiff(sp_list_by_day[[a]], sp_list_by_day[[a-1]]))
      }
   
     if (a == 1){
      cum_list <- sp_list_by_day[[a]]} else {cum_list <- c(cum_sp_by_day[[a-1]],sp_list_by_day[[a]])}
    #now get rid of duplicates
    cum_list <- unique(cum_list)
    cum_list
    #add species to cumulative list of species by day
    cum_sp_by_day[[a]] <- cum_list
    
    
    cum_num_species <- n_distinct(cum_list)
    
    temp_output <- as.data.frame(list(my_deployment, my_date, num_sp_today, num_sp_shared_yesterday, num_new_sp, cum_num_species))
   
      names(temp_output) <- names(species_accum)

    species_accum <- rbind(species_accum, temp_output)

  }
  
  #now add the species list by day for a single deployment to the list of species lists by deployment
  sp_list_by_deployment[[i]] <- sp_list_by_day
}
```

Take a look:

```{r}
glimpse(species_accum)

treatment <- deploy |> select(deployment_num, AM_housing, paired, paired_site, deploy_len)

species_accum <- left_join(species_accum, treatment, by =  join_by("deployment_num" == "deployment_num"))
```

#### Paired Accumulation graphs
```{r}
#| fig-width: 12
#| out-width: "100%"
paired_accum <- species_accum |> filter(paired == "y")

ggplot(paired_accum, 
aes(date, cum_num_species, group = deployment_num, color = AM_housing))+
  geom_line(linewidth = 1)+
  #geom_smooth(se = F)+
  facet_wrap(~paired_site)+
  labs(x = "Date", y = "Cumulative number of species")+
  labs(color = "AM Housing")+
  scale_colour_viridis_d()+
  theme_minimal(base_size = 14)
```

#### Overall Mean Accumulation by Treatment


```{r}
all_accum <- species_accum |> 
    mutate(Round = if_else(deploy_len <= 21, 
    true = "Round 1", false = "Round 2")) |> 
    group_by(deployment_num) |> 
    mutate(index = row_number()) |> 
    ungroup()

ggplot(all_accum, 
aes(index, cum_num_species, group = deployment_num, color = AM_housing))+
  geom_line()+
  #geom_smooth(se = F)+
  facet_wrap(~Round)+
  labs(x = "Day deployed", y = "Cumulative number of species")+
  labs(color = "AM Housing")+
  scale_colour_viridis_d()+
  theme_minimal(base_size = 12)
```

## Code Appendix

```{r ref.label=knitr::all_labels()}
#| echo: true
#| eval: false
```
