---
title: "Exploring data"
format: html
---
## Introduction

This document contains the final analysis for the baggie case paper. The starting point will be the outline of the result section that I have already written, and as analysis reveals insight/new questions, I will add/subtract from there. I will start by calculating some general statistics about the data and the sampling effort. Then, I will look at paired data subset, first by creating a data frame that will be used in all the possible analyses for those data. After, I will continue with the analysis for the whole project. 

## Set up

Let's load in the necessary packages and the merged data file (containing both deployment data and BirdNET output).

```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(pander)
library(sf)
library(spmodel)
merged <- read_csv(here("data/merged_deployment_birdnet.csv"))
birdnet <- read_csv(here("data/birdnet_results.csv"))
deploy <- read_csv("data/Updated_baggie_case_deployments.csv")
history <- read_csv("data/acoustic_device_deployments.csv")
```

Let's take a look at the structure:

```{r}
glimpse(merged)
```


## Summary Statistics

There are some variables to add to the data frame:

 - record_len: stores the length of actual recording (using the deployment date and last recording date)

 - battery_life: stores the expected battery life based on the config file (same for every AM)

 - *gb_full: amount of time expected to fill the sd card based on card size and config file*
     - *Would have added sd_full variable but the battery seems like the limiting factor (regardless of the 64 GB or 256 GB cards)*

```{r}
# save recording length and battery life as new variables
merged <- merged |> mutate(dropoff_date = dmy(dropoff_date),
last_record = mdy(last_record),
record_len = last_record - dropoff_date,
battery_life = rep(2500/60, nrow(merged)), # 2500 mAh with a daily usage of 60 mAh based on config
gb_per_d = rep(1.162, nrow(merged))) |> # GB used per day based on config
filter(deployment_num != "AM0100") # the known AM with user error

# add these same variables to the deployment data (also change pick up data type):
deploy <- deploy |> mutate(dropoff_date = dmy(dropoff_date),
last_record = mdy(last_record),
record_len = last_record - dropoff_date,
battery_life = rep(2500/60, nrow(deploy)),
gb_per_d = rep(1.162, nrow(deploy)),
pickup_date = dmy(pickup_date))|> 
filter(deployment_num != "AM0100") 

glimpse(merged)
glimpse(deploy)
```


Total nights of deployment and nights of recording:

```{r}
deploy |> summarise(tot_deploy = sum(pickup_date - dropoff_date, na.rm = T),
tot_rec = sum(record_len, na.rm = T)) |> pander()
```

Total ids and total species richness:

```{r}
birdnet |> summarise(tot_ids = n(),
tot_sp = n_distinct(species)) |> pander()
```

Total identifications and species richness per AM:

```{r}
merged |> group_by(deployment_num) |> 
    summarise(num_ids = sum(!is.na(probability)),
    num_sp = n_distinct(species, na.rm = T)) |> 
        pander()
```

#### Baggie vs Case Comparisons

Total identifications and species richness:

```{r}
merged |> group_by(AM_housing) |>
    summarise(total_ids = sum(!is.na(probability)),
    total_sp = n_distinct(species, na.rm = T)) |> 
        pander() 
```

Mean and sd:

```{r}
merged |> group_by(AM_housing, deployment_num) |> 
    summarise(total_ids = sum(!is.na(probability)),
    total_sp = n_distinct(species, na.rm = T)) |> 
        group_by(AM_housing) |> 
        summarise(mean_ids = mean(total_ids),
        sd_ids = sd(total_ids),
        mean_sp = mean(total_sp),
        sd_sp = sd(total_sp)) |> 
            pander()
```

## Paired Analysis

Here the analysis will focus on the subset of deployments that included two AMs at the same site.

Let's create the data frame needed for the paired analysis:

```{r}
# create a dataset that has the species richness and total IDs for each AM
rich_and_ids <- merged |> filter(paired == "y") |> 
    group_by(deployment_num) |> 
    summarise( id_count = sum(!is.na(probability)),
    species_count = n_distinct(species, na.rm = T))

# merge the deployment 
paired <- deploy |> filter(paired == "y") |> 
    select(deployment_num, AM_housing, paired_site, record_len) |> 
    left_join(rich_and_ids, join_by("deployment_num" == "deployment_num")) |> 
    mutate(id_count_adj = id_count/as.numeric(record_len),
    species_count_adj = species_count/as.numeric(record_len)) |> # divides totals by the duration of recording
    mutate(across(c(id_count_adj, species_count_adj), ~replace_na(., 0)))|> 
    pivot_wider(id_cols = paired_site, names_from = AM_housing,
    values_from = c(id_count_adj, species_count_adj)) |> 
        mutate(diff_detect = id_count_adj_Case - id_count_adj_Ziplock,
        diff_SR = species_count_adj_Case - species_count_adj_Ziplock)
```

#### Statistical test

Since we have a low number of replicates, we will use a Wilcoxon Signed Rank Test

```{r}
wilcox.test(paired$id_count_adj_Case, paired$id_count_adj_Ziplock)

wilcox.test(paired$species_count_adj_Case, paired$species_count_adj_Ziplock)
```

Not significant, but there still might be a trend

#### Plot

species richness
```{r}
# data for plot
paired_plot <- deploy |> filter(paired == "y") |> 
    select(deployment_num, AM_housing, paired_site, record_len) |> 
    left_join(rich_and_ids, join_by("deployment_num" == "deployment_num")) |> 
    mutate(id_count_adj = id_count/as.numeric(record_len),
    species_count_adj = species_count/as.numeric(record_len)) |> 
    mutate(across(c(id_count_adj, species_count_adj), ~replace_na(., 0)))

# species richness
ggplot(paired_plot, aes(AM_housing, species_count, group = paired_site))+
    geom_point(aes(color = AM_housing))+
    geom_line()+
    theme_minimal()

# total ids
ggplot(paired_plot, aes(AM_housing, id_count, group = paired_site))+
    geom_point(aes(color = AM_housing))+
    geom_line()+
    theme_minimal()
```

## Durability Analysis

Due to the estimation by the configuration app based on our config file, we know that the batteries will last around 41.67 days and we used 1.162 GB on the SD card per day. Since the SD cards were either 64 GB or 256 GB, the batteries are the clear limiting factor on how long we would predict the AM to last. Therefore, for our durability analysis, our response variable will be the percentage of days that the AM lasted based on our expected battery life.

Let's add this variable to the deployment dataset:


```{r}
durability <- deploy |> 
    mutate(record_len = 
    replace(record_len, deployment_num == "AM0057", 0), # AM did not record at all (coded as 0)
    percentage = record_len/battery_life)
```

#### Basic statistics

```{r}
durability |> 
    group_by(AM_housing) |> 
    summarise(mean_percent = mean(percentage),
    sd_percent = sd(percentage)) |> 
        pander()
```

What if we remove the failed AM?

```{r}
durability |>
    filter(deployment_num != "AM0057") |> 
    group_by(AM_housing) |> 
    summarise(mean_percent = mean(percentage),
    sd_percent = sd(percentage)) |>
        pander()
```

The cases lasted for a smaller percentage of the expected recording duration on average!

However, there is less variation in recording duration in the case group (especially after removing the failed AM).

Let's see a visual of these results:

Side-by-side boxplots:

```{r}
ggplot(durability, aes(AM_housing, percentage))+
    geom_boxplot()+
    geom_point()+
    theme_minimal()
```

Very intersting! 

#### Historical Deployments

Perhaps we are looking at the wrong variable. Clearly housing does not seem to affect durability much, but previous deployments might be a much stronger predictor of durability. 

Using the complete deployment history of each audiomoth we can see if the total deployment days influences the 'recording length' of the first and second round of deployments in this experiment.


Getting rid of AM0057 for this analysis:

```{r}
deploy_cleaned <- deploy |> filter(deployment_num != "AM0057")
```

We need to filter out all the deployments after the baggie/case experiment:
```{r}
# this code is grabbing all the 'first round' deployments 
# and finding the earliest pick up date
# this way we can combine it with all the historical deployments and look
round_1_end <- deploy_cleaned |> 
    group_by(recorder_lab_id_num) |> 
    arrange(pickup_date) |> 
    filter(n() >= 2) |> 
    slice(1) |> 
    ungroup() |> 
    filter(pickup_date == max(pickup_date)) |> 
    slice(1)

# for this time frame we are looking for first deployment of the baggie/case experiment
# this way we can look at the effect of the historical deployments on the functional
# length of the first baggie/case deployment
first_deploy <- deploy_cleaned |> 
    filter(dropoff_date == min(dropoff_date)) |> 
    slice(1)

# we also need the list of AMs that had two deployments during the experiment
two_deploy <- deploy_cleaned |> 
    group_by(recorder_lab_id_num) |> 
    filter(n() >= 2)

history_2_deploy <- history |> 
    select(deployment_num, am_recorder_lab_id_num, dropoff_date, pickup_date, AM_housing) |> 
    mutate(dropoff_date = lubridate::dmy(dropoff_date),
    pickup_date = lubridate::dmy(pickup_date)) |> 
    filter(pickup_date <= round_1_end$pickup_date) |> 
    filter(am_recorder_lab_id_num %in% two_deploy$recorder_lab_id_num)

history_before <- history |> 
    select(deployment_num, am_recorder_lab_id_num, dropoff_date, pickup_date, AM_housing) |> 
    mutate(dropoff_date = lubridate::dmy(dropoff_date),
    pickup_date = lubridate::dmy(pickup_date)) |> 
    filter(pickup_date < first_deploy$pickup_date) |> 
    filter(am_recorder_lab_id_num %in% deploy_cleaned$recorder_lab_id_num)
    # only look at AMs that we deployed for the experiment
```

Regression with total number of days of deployments and recording length on the last deployment:

```{r}
record_len <- durability |> arrange(dropoff_date) |> 
    group_by(recorder_lab_id_num) |> 
    summarise(first_rec_len = first(record_len),
    last_rec_len = last(record_len))

after_first_deploy <- history_2_deploy |> group_by(am_recorder_lab_id_num) |> 
    summarise(tot_days = sum(pickup_date - dropoff_date),
    n_deploy = n()) |> 
    ungroup() |> 
    left_join(record_len, join_by(am_recorder_lab_id_num == recorder_lab_id_num))

before_exp <- history_before |> 
    filter(dropoff_date < first_deploy$dropoff_date) |>
    group_by(am_recorder_lab_id_num) |> 
    summarise(tot_days = sum(pickup_date - dropoff_date),
    n_deploy = n()) |> 
    ungroup() |> 
    left_join(record_len, join_by(am_recorder_lab_id_num == recorder_lab_id_num))

# second deployment model
second_deploy_mod <- lm(as.numeric(last_rec_len)~as.numeric(tot_days), data = after_first_deploy)
summary(second_deploy_mod)

# first deployment model
first_deploy_mod <- lm(as.numeric(first_rec_len)~as.numeric(tot_days), data = before_exp)
summary(first_deploy_mod)

# second deployment data:
# plot total number of days deployed
ggplot(after_first_deploy, aes(tot_days, last_rec_len))+
    geom_point()+
    geom_smooth(se = F)+
    theme_minimal()

# plot number of deployments
ggplot(after_first_deploy, aes(n_deploy, last_rec_len))+
    geom_point()+
    geom_smooth(se = F)+
    theme_minimal()

# first deployment data:
# plot total number of days deployed
ggplot(before_exp, aes(tot_days, first_rec_len))+
    geom_point()+
    geom_smooth(se = F)+
    theme_minimal()

# plot number of deployments
ggplot(before_exp, aes(n_deploy, first_rec_len))+
    geom_point()+
    geom_smooth(se = F)+
    theme_minimal()
```

It doesn't seem like there's much of an effect from historical deployments. In general, both baggie/case and historical deployments don't seem to have much of an effect on the performance/durability of the AMs in the field.


## All Sites

Now for the big finale... we will look at the the entire data set, both paired and upaired, to see if either housing group is significantly better at detecting more species (species richness) or just number of detections.

We have two candidate models: 

1. Multilevel Model: $SR = \beta_0 + \beta_1case\_type + \epsilon$
 - where the random effect is for paired sites
 - assumption is that all unpaired sites are independent

2. $SR = \beta_0 + \beta_1case\_type + \epsilon + \tau$
 - where the random effect is for paired sites
 - $\tau$ is a spatial dependent term with some covariance structure: 
  * $Cov(\tau_i, \tau_j) = \sigma^2 exp(-d_{ij} / \phi)$
 - no point is considered spatially independent

We will use `spmodel` for both models so that we can compare them with the same AIC/BIC.

#### Multilevel Model for SR

First we need to manipulate the the deployment data to have a unique ID for each paried site and re-join with the birdnet output.

```{r}
deploy_mod_df <- deploy_cleaned |> mutate(pair_id = if_else(paired == 'y',
paste(pmin(deployment_num, paired_with), 
pmax(deployment_num, paired_with), 
sep = "_"), deployment_num)) # if AM is not paired, it just gets its deploy number

SR_count_df <- merged |> group_by(deployment_num) |> 
    summarise(count = n(),
    sr = n_distinct(species))

full_mod_df <- left_join(deploy_mod_df, SR_count_df, join_by(deployment_num == deployment_num)) |> 
    mutate(count_adj = count/as.numeric(record_len),
    sr_adj = sr/as.numeric(record_len))
```

Convert this data frame into an sf object using the NAD83 projection (EPSG: 4269):

```{r}
full_mod_sf <- st_as_sf(full_mod_df, coords = c("longitude", "latitude"), crs = 4269)
```

Now we can use splm() to create the model:

```{r}
mixed_mod <- splm(sr_adj ~ AM_housing, data = full_mod_sf, 
random = ~ 1 | pair_id, spcov_type = "none")
summary(mixed_mod)
```

No significant p-value now that we account for recording length

#### Spatial Model for SR

Now we can try the spatial model:

```{r}
spatial_mod <- splm(sr_adj ~ AM_housing, data = full_mod_sf, 
spcov_type = "exponential", # added spatial covariance term
random = ~ 1 | pair_id)
summary(spatial_mod)
```

Same deal.

 Comparison:

```{r}
glances(mixed_mod, spatial_mod)
```

Looks like the the standard mixed model did better and standardizing for recording length seems to have increased the near-significant p-value for the mixed model. 

#### Plots

First lets look at a side-by-side boxplot for all the sites

```{r}
ggplot(full_mod_df, aes(AM_housing, sr))+
    geom_boxplot()+
    theme_minimal()
```



