---
title: "Exploring data"
format: html
---
## Introduction

This document contains the final analysis for the baggie case paper. The starting point will be the outline of the result section that I have already written, and as analysis reveals insight/new questions, I will add/subtract from there. I will start by calculating some general statistics about the data and the sampling effort. Then, I will look at paired data subset, first by creating a data frame that will be used in all the possible analyses for those data. After, I will continue with the analysis for the whole project. 

## Set up

Let's load in the necessary packages and the merged data file (containing both deployment data and BirdNET output).

```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(pander)
merged <- read_csv(here("data/merged_deployment_birdnet.csv"))
birdnet <- read_csv(here("data/birdnet_results.csv"))
deploy <- read_csv("data/Updated_baggie_case_deployments.csv")
```

Let's take a look at the structure:

```{r}
glimpse(merged)
```


## Summary Statistics

There are some variables to add to the data frame:

 - record_len: stores the length of actual recording (using the deployment date and last recording date)

 - battery_life: stores the expected battery life based on the config file (same for every AM)

 - *sd_full: amount of time expected to fill the sd card based on card size and config file*

```{r}
# save recording length and battery life as new variables
merged <- merged |> mutate(dropoff_date = dmy(dropoff_date),
last_record = mdy(last_record),
record_len = last_record - dropoff_date,
battery_life = rep(2500/60, nrow(merged)), # 2500 mAh with a daily usage of 60 mAh based on config
files_per_d = rep(1.162, nrow(merged))) |> # GB used per day based on config
filter(deployment_num != "AM0100") # the known AM with user error

# add these same variables to the deployment data (also change pick up data type):
deploy <- deploy |> mutate(dropoff_date = dmy(dropoff_date),
last_record = mdy(last_record),
record_len = last_record - dropoff_date,
battery_life = rep(2500/60, nrow(deploy)),
files_per_d = rep(1.162, nrow(merged)),
pickup_date = dmy(pickup_date))

glimpse(merged)
glimpse(deploy)
```

Total nights of deployment and nights of recording:

```{r}
deploy |> summarise(tot_deploy = sum(pickup_date - dropoff_date, na.rm = T),
tot_rec = sum(record_len, na.rm = T)) |> pander()
```

Total ids and total species richness:

```{r}
birdnet |> summarise(tot_ids = n(),
tot_sp = n_distinct(species)) |> pander()
```

Total identifications and species richness per AM:

```{r}
merged |> group_by(deployment_num) |> 
    summarise(num_ids = sum(!is.na(probability)),
    num_sp = n_distinct(species, na.rm = T)) |> 
        pander()
```

#### Baggie vs Case Comparisons

Total identifications and species richness:

```{r}
merged |> group_by(AM_housing) |>
    summarise(total_ids = sum(!is.na(probability)),
    total_sp = n_distinct(species, na.rm = T)) |> 
        pander() 
```

Mean and sd:

```{r}
merged |> group_by(AM_housing, deployment_num) |> 
    summarise(total_ids = sum(!is.na(probability)),
    total_sp = n_distinct(species, na.rm = T)) |> 
        group_by(AM_housing) |> 
        summarise(mean_ids = mean(total_ids),
        sd_ids = sd(total_ids),
        mean_sp = mean(total_sp),
        sd_sp = sd(total_sp))
```




