---
title: "Figures and Tables"
author: "Aidan Fauth"
execute: 
  warning: false
  echo: false
format: 
  docx:
    embed-resources: true
---

This document will contain al the figures found in my SYE paper.

```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(pander)
library(sf)
library(spmodel)
merged <- read_csv(here("data/merged_deployment_birdnet.csv"))
birdnet <- read_csv(here("data/birdnet_results.csv"))
deploy <- read_csv(here("data/Updated_baggie_case_deployments.csv"))
```

Clean up data frames:

```{r}
# save recording length and battery life as new variables
merged <- merged |> mutate(dropoff_date = dmy(dropoff_date),
last_record = mdy(last_record),
record_len = last_record - dropoff_date,
battery_life = rep(2500/60, nrow(merged)), # 2500 mAh with a daily usage of 60 mAh based on config
gb_per_d = rep(1.162, nrow(merged))) |> # GB used per day based on config
filter(deployment_num != "AM0100" & deployment_num != "AM0057") # the known AM with user error

# add these same variables to the deployment data (also change pick up data type):
deploy <- deploy |> mutate(dropoff_date = dmy(dropoff_date),
last_record = mdy(last_record),
record_len = last_record - dropoff_date,
battery_life = rep(2500/60, nrow(deploy)),
gb_per_d = rep(1.162, nrow(deploy)),
pickup_date = dmy(pickup_date),
deploy_len = pickup_date - dropoff_date)|> 
filter(deployment_num != "AM0100" & deployment_num != "AM0057") 
```

No anthroprogenic noise:

```{r}
anthro <- c("Fireworks_Fireworks", "Siren_Siren", "Engine_Engine", "Gun_Gun", 
  "Human vocal_Human vocal", "Human non-vocal_Human non-vocal", "Dog_Dog")

cleaned <- merged |> filter(!(species %in% anthro))
```

## Basic Summary Stats

Total nights of deployment and nights of recording:

```{r}
deploy |> summarise(tot_deploy = sum(pickup_date - dropoff_date, na.rm = T),
tot_rec = sum(record_len, na.rm = T)) |> pander()
```

Total ids and total species richness:

```{r}
birdnet |> summarise(tot_ids = n(),
tot_sp = n_distinct(species)) |> pander()
```

Total identifications and species richness per AM:

```{r}
merged |> group_by(deployment_num) |> 
    summarise(num_ids = sum(!is.na(probability)),
    num_sp = n_distinct(species, na.rm = T)) |> 
        pander()
```

#### Baggie vs Case Comparisons

Total identifications and species richness:

```{r}
merged |> group_by(AM_housing) |>
    summarise(total_ids = sum(!is.na(probability)),
    total_sp = n_distinct(species, na.rm = T)) |> 
        pander() 

cleaned |> group_by(AM_housing) |>
    summarise(total_ids = sum(!is.na(probability)),
    total_sp = n_distinct(species, na.rm = T)) |> 
        pander() 
```

Mean and sd:

```{r}
merged |> group_by(AM_housing, deployment_num) |> 
    summarise(total_ids = sum(!is.na(probability)),
    total_sp = n_distinct(species, na.rm = T)) |> 
        group_by(AM_housing) |> 
        summarise(mean_ids = mean(total_ids),
        sd_ids = sd(total_ids),
        mean_sp = mean(total_sp),
        sd_sp = sd(total_sp)) |> 
            pander()

cleaned |> group_by(AM_housing, deployment_num) |> 
    summarise(total_ids = sum(!is.na(probability)),
    total_sp = n_distinct(species, na.rm = T)) |> 
        group_by(AM_housing) |> 
        summarise(mean_ids = mean(total_ids),
        sd_ids = sd(total_ids),
        mean_sp = mean(total_sp),
        sd_sp = sd(total_sp)) |> 
            pander()
```



#### Baggie & Case Species Comparison

```{r}
Species <- unique(merged$species)

case_species <- merged |> filter(AM_housing == "Case")

zip_species <- merged |> filter(AM_housing == "Ziplock")

Case <- c()
Ziplock <- c()

for (i in Species){
    if (i %in% case_species$species & i %in% zip_species$species){
        Case <- c(Case, "X")
        Ziplock <- c(Ziplock, "X")
    } else if (i %in% case_species$species){
        Case <- c(Case, "X")
        Ziplock <- c(Ziplock, "")
    }else {
        Case <- c(Case, "")
        Ziplock <- c(Ziplock, "X")
    }

}

b_c_list <- tibble(Species, Case, Ziplock)

b_c_list |> arrange(Species)|> 
    separate(Species, into = c("Speices", "Common Name"), sep = "_")|> 
    pander()
```


## Durability

```{r}
durability <- deploy |> 
    mutate(percentage = record_len/battery_life,
    deploy_len = pickup_date - dropoff_date)

durability_test <- durability |> 
    select(deployment_num, AM_housing, deploy_len, record_len, percentage) |> 
    filter(deploy_len > 21)

durability |> group_by(AM_housing) |> summarise(num = sum(record_len),
mean = mean(record_len),
sd = sd(record_len)) |> pander()

durability_test |> group_by(AM_housing) |> summarise(num = sum(record_len),
mean = mean(record_len),
sd = sd(record_len)) |> pander()

ggplot(durability_test, aes(AM_housing, record_len))+
    geom_boxplot()+
    geom_jitter(width = 0.03, height = 0)+
    geom_hline(yintercept = 41.67, linetype = "dashed", color = "red")+
    scale_y_continuous(limits = c(0, NA))+
    labs(x = "Housing Type",
    y = "Recording length (days)")+
    theme_minimal()
```

## Paired

### IDs

```{r}
# create a dataset that has the species richness and total IDs for each AM
ids <- merged |> 
    group_by(deployment_num, day(date_time)) |> 
    summarise( id_day = sum(!is.na(probability))) |> 
    ungroup() |> 
    group_by(deployment_num) |> 
    summarise(id_mean = mean(id_day)) |> 
    ungroup()

# merge the deployment 
paired_ids <- deploy |> filter(paired == "y") |> 
    select(deployment_num, AM_housing, paired_site, record_len) |> 
    left_join(ids, join_by("deployment_num" == "deployment_num")) |> 
    pivot_wider(id_cols = paired_site, names_from = AM_housing,
    values_from = id_mean) |> 
        mutate(diff_detect = Case - Ziplock)
```


#### Plot

species richness
```{r}
# data for plot
paired_plot_ids <- deploy |> filter(paired == "y") |> 
    select(deployment_num, AM_housing, paired_site, record_len) |> 
    left_join(ids, join_by("deployment_num" == "deployment_num"))

# total ids
ggplot(paired_plot_ids, aes(AM_housing, id_mean, group = paired_site))+
    geom_point(aes(color = AM_housing))+
    geom_line()+
    labs(x = "Housing Type",
        y = "Average Detections per day",
        color = "Housing Type")+
    theme_minimal()+
    theme(legend.position = "none")
```


### Species Richness

Remove anthroprogenic noises

```{r}
# create a dataset that has the species richness and total IDs for each AM
rich_cleaned <- cleaned |>
    group_by(deployment_num, day(date_time)) |> 
    summarise(sr_day = n_distinct(species, na.rm = T)) |> 
    ungroup() |> 
    group_by(deployment_num) |> 
    summarise(sr_mean = mean(sr_day)) |> 
    ungroup()

# merge the deployment 
paired_sr <- deploy |> filter(paired == "y") |> 
    select(deployment_num, AM_housing, paired_site, record_len) |> 
    left_join(rich_cleaned, join_by("deployment_num" == "deployment_num")) |> 
    pivot_wider(id_cols = paired_site, names_from = AM_housing,
    values_from = sr_mean) |> 
        mutate(diff_SR = Case - Ziplock)
```


#### Plot

```{r}
paired_plot_sr <- deploy |> filter(paired == "y") |> 
    select(deployment_num, AM_housing, paired_site, record_len) |> 
    left_join(rich_cleaned, join_by("deployment_num" == "deployment_num"))

# species richness
ggplot(paired_plot_sr, aes(AM_housing, sr_mean, group = paired_site))+
    geom_point(aes(color = AM_housing))+
    geom_line()+
    labs(x = "Housing Type",
        y = "Average SR per day",
        color = "Housing Type")+
    theme_minimal()+
    theme(legend.position = "none")
```


## Model


```{r}
overall_plot <- left_join(deploy, rich_cleaned, join_by(deployment_num == deployment_num)) |> 
    mutate(
    paired = fct_recode(paired, "paired" = "y", "not paired" = "n"))

ggplot(overall_plot, aes(AM_housing, sr_mean))+
    geom_boxplot(color = "black")+
    geom_jitter(width = 0.04, height = 0, aes(color = paired))+
    scale_color_manual(values = c("paired" = "red", "not paired" = "black"))+
    labs(x = "Housing Type",
    y = "Average species per day",
    color = "Site Type")+
    theme_minimal()

id_plot <- left_join(deploy, ids, join_by(deployment_num == deployment_num)) |> 
    mutate(
    paired = fct_recode(paired, "paired" = "y", "not paired" = "n"))

ggplot(id_plot, aes(AM_housing, id_mean))+
    geom_boxplot(color = "black")+
    geom_jitter(width = 0.04, height = 0, aes(color = paired))+
    scale_color_manual(values = c("paired" = "red", "not paired" = "black"))+
    labs(x = "Housing Type",
    y = "Average species IDs per day",
    color = "Site Type")+
    theme_minimal()
```



## Species Accumulation

```{r}
species_accum <- tibble(
  deployment_num = character(),
  date = POSIXct(),
  num_sp_today = integer(),
  num_sp_shared_yesterday = integer(),
  num_new_sp = integer(),
  cum_num_species = integer()
)
```

```{r}
cleaned <- cleaned |> mutate(date = date(date_time))

deployments <- unique(cleaned$deployment_num)
#empty list to hold lists of species per detector
#create list using deployments so that each list element is named to a particular deployment

sp_list_by_deployment <- sapply(deployments, function(x) NULL)

#second empty list to hold lists of cumulative species per detector
cum_sp_list_by_deployment <- sapply(deployments, function(x) NULL)

#start loop
for(i in 1:length(deployments)){
  my_deployment <- deployments[i]

  temp_df <- cleaned|> filter(deployment_num == my_deployment) |> arrange(deployment_num, date_time)
  
  #second loop to isolate days
  days <- unique(temp_df$date)
  #empty list to hold species lists
  sp_list_by_day <- vector(mode = "list", length = length(days))
  
  #second empty list to hold cumulative species lists
  cum_sp_by_day <- vector(mode = "list", length = length(days))
  
  for(a in 1:length(days)){
    my_date <- days[a]
    
    day_df <- temp_df |> filter(date == my_date)
    num_sp_today <- n_distinct(day_df$species)
    
    sp_list_by_day[[a]] <- unique(day_df$species)
    if (a == 1){
      num_sp_shared_yesterday <- 0} else {
      num_sp_shared_yesterday <- n_distinct(intersect(sp_list_by_day[[a-1]], sp_list_by_day[[a]]))
      }
    if (a == 1){
      num_new_sp <- num_sp_today } else {
        num_new_sp <- length(setdiff(sp_list_by_day[[a]], sp_list_by_day[[a-1]]))
      }
   
     if (a == 1){
      cum_list <- sp_list_by_day[[a]]} else {cum_list <- c(cum_sp_by_day[[a-1]],sp_list_by_day[[a]])}
    #now get rid of duplicates
    cum_list <- unique(cum_list)
    cum_list
    #add species to cumulative list of species by day
    cum_sp_by_day[[a]] <- cum_list
    
    
    cum_num_species <- n_distinct(cum_list)
    
    temp_output <- as.data.frame(list(my_deployment, my_date, num_sp_today, num_sp_shared_yesterday, num_new_sp, cum_num_species))
   
      names(temp_output) <- names(species_accum)

    species_accum <- rbind(species_accum, temp_output)

  }
  
  #now add the species list by day for a single deployment to the list of species lists by deployment
  sp_list_by_deployment[[i]] <- sp_list_by_day
}
```

Take a look:

```{r}
glimpse(species_accum)

treatment <- deploy |> select(deployment_num, AM_housing, paired, paired_site, deploy_len)

species_accum <- left_join(species_accum, treatment, by =  join_by("deployment_num" == "deployment_num"))
```

#### Paired Accumulation graphs
```{r}
#| fig-width: 12
#| out-width: "100%"
paired_accum <- species_accum |> filter(paired == "y")

ggplot(paired_accum, 
aes(date, cum_num_species, group = deployment_num, color = AM_housing))+
  geom_line(linewidth = 1)+
  #geom_smooth(se = F)+
  facet_wrap(~paired_site)+
  labs(x = "Date", y = "Cumulative number of species")+
  labs(color = "AM Housing")+
  theme_minimal(base_size = 14)
```

#### Overall Mean Accumulation by Treatment


```{r}
all_accum <- species_accum |> 
    mutate(Round = if_else(deploy_len <= 21, 
    true = "Round 1", false = "Round 2")) |> 
    group_by(deployment_num) |> 
    mutate(index = row_number()) |> 
    ungroup()

ggplot(all_accum, 
aes(index, cum_num_species, group = deployment_num, color = AM_housing))+
  geom_line()+
  #geom_smooth(se = F)+
  facet_wrap(~Round)+
  labs(x = "Day deployed", y = "Cumulative number of species")+
  labs(color = "AM Housing")+
  theme_minimal(base_size = 12)
```

## Models

#### Multilevel Model for SR

First we need to manipulate the the deployment data to have a unique ID for each paried site and re-join with the birdnet output.

```{r}
deploy_mod_df <- deploy |> mutate(pair_id = if_else(paired == 'y',
paste(pmin(deployment_num, paired_with), 
pmax(deployment_num, paired_with), 
sep = "_"), deployment_num)) |> 
mutate(deploy_len = pickup_date - dropoff_date,
Round = if_else(deploy_len <= 21, 
    true = "Round 1", false = "Round 2"))

sr_mod_df <- left_join(deploy_mod_df, rich_cleaned, 
join_by(deployment_num == deployment_num))
```

Convert this data frame into an sf object using the NAD83 projection (EPSG: 5070):

```{r}
#sr_mod_sf <- st_as_sf(sr_mod_df, coords = c("longitude", "latitude"), crs = 5070)

sr_mod_sf <- st_as_sf(sr_mod_df, coords = c("longitude", "latitude"), crs = 4326)

sr_mod_sf <- st_transform(sr_mod_sf, crs = 5070)
```

Now we can use splm() to create the model:

```{r}
mixed_mod <- splm(sr_mean ~ AM_housing + Round, data = sr_mod_sf, 
random = ~ 1 | pair_id, spcov_type = "none")
summary(mixed_mod)
```

significant p-value now that we account for round and average per day!

```{r}
tidy(mixed_mod) |> pander()
```

#### Spatial Model for SR

Now we can try the spatial model:

```{r}
spatial_mod <- splm(sr_mean ~ AM_housing + Round, data = sr_mod_sf, 
spcov_type = "exponential")
summary(spatial_mod)
```

Same deal.

 Comparison:

$$ y = \beta_0 + \beta_1(AM Housing) + \beta_2(Round) + \epsilon_{site} 
$$

$$ Y = \beta_0 + \beta_1(AM Housing) + \beta_2(Round) + \tau 
$$

```{r}
glances(mixed_mod, spatial_mod) |> pander()
```

Looks like the the standard mixed model did better and standardizing for recording length seems to have increased the near-significant p-value for the mixed model. 

#### Mixed model for IDs

```{r}
deploy_mod_df <- deploy |> mutate(pair_id = if_else(paired == 'y',
paste(pmin(deployment_num, paired_with), 
pmax(deployment_num, paired_with), 
sep = "_"), deployment_num)) |> 
mutate(deploy_len = pickup_date - dropoff_date,
Round = if_else(deploy_len <= 21, 
    true = "Round 1", false = "Round 2"))

id_mod_df <- left_join(deploy_mod_df, ids, 
join_by(deployment_num == deployment_num))
```

Convert this data frame into an sf object using the NAD83 projection (EPSG: 5070):

```{r}
id_mod_sf <- st_as_sf(id_mod_df, coords = c("longitude", "latitude"), crs = 5070)
```

Now we can use splm() to create the model:

```{r}
mixed_id_mod <- splm(id_mean ~ AM_housing + Round, data = id_mod_sf, 
random = ~ 1 | pair_id, spcov_type = "none")
summary(mixed_mod)
```

```{r}
tidy(mixed_id_mod) |> pander()
```


