---
title: "Preliminary Analysis"
format: html
---

## Goal:

This document looks to continue to explore/vizualize the results of the baggie/case experiment and look at some preliminary analyses.

## Set up

```{r}
rm(list = ls())
library(tidyverse)
library(here)
```

## Load in Data

```{r}
birdnet <- read_csv(here("data/birdnet_results.csv"))
deploy <- read_csv(here("data/Updated_baggie_case_deployments.csv"))
```

#### Merge dataframes

```{r}
merged <- left_join(deploy, birdnet, join_by("deployment_num" == "id"))
glimpse(merged)
```

Note: there will be some rows that only have deployment info because two of the audio moths failed

We know that one of the deployments failed because of user error.

We can remove that one because it does not have to do with durability:

```{r}
cleaned <- merged |> filter(deployment_num != "AM0100") # the known AM with user error
```

#### Calculate the duration variable and battery-life

```{r}
cleaned <- cleaned |> 
    mutate(last_record = mdy(last_record),
    dropoff_date = dmy(dropoff_date),
    pickup_date = dmy(pickup_date)) |> 
    group_by(deployment_num) |> 
    mutate(duration = last_record - dropoff_date,
    deploymet_len = pickup_date - dropoff_date) |> 
    ungroup() |> 
    mutate(E_battery_life = rep(2500/60, nrow(cleaned)),
    diff_E_bat = E_battery_life - duration, 
    duration = replace(duration, deployment_num == "AM0057", 0)) # the recorder that failed without known cause
```

## Paired data:

```{r}
rich_and_ids <- cleaned |> filter(paired == "y") |> 
    group_by(deployment_num) |> 
    summarise(id_count = n(), 
    species_count = n_distinct(species))

deploy_cleaned <- deploy |> 
    filter(deployment_num != "AM0100") |> 
    mutate(last_record = mdy(last_record),
    dropoff_date = dmy(dropoff_date),
    pickup_date = dmy(pickup_date)) |> 
    group_by(deployment_num) |> 
    mutate(duration = last_record - dropoff_date,
    deployment_len = pickup_date - dropoff_date) |> 
    ungroup() |> 
    mutate(duration = replace(duration, deployment_num == "AM0057", 0)) |> 
    ungroup() 
    
    
paired <- deploy_cleaned |> filter(paired == "y") |> 
    select(deployment_num, AM_housing, paired_site) |> 
    left_join(rich_and_ids, join_by("deployment_num" == "deployment_num")) |> 
    mutate(across(c(id_count, species_count), ~replace_na(., 0)))|> 
    pivot_wider(id_cols = paired_site, names_from = AM_housing,
    values_from = c(id_count, species_count)) |> 
        mutate(diff_detect = id_count_Case - id_count_Ziplock,
        diff_SR = species_count_Case - species_count_Ziplock)
        
```


#### Statistical test

Since we have a low number of replicates, we will use a Wilcoxon Signed Rank Test

```{r}
wilcox.test(paired$id_count_Case, paired$id_count_Ziplock)

wilcox.test(paired$species_count_Case, paired$species_count_Ziplock)
```

Not significant, but there still might be a trend

### Plot

species richness
```{r}
# data for plot
paired_plot <- deploy_cleaned |> filter(paired == "y") |> 
    select(deployment_num, AM_housing, paired_site) |> 
    left_join(rich_and_ids, join_by("deployment_num" == "deployment_num")) |> 
    mutate(across(c(id_count, species_count), ~replace_na(., 0)))

ggplot(paired_plot, aes(AM_housing, species_count, group = paired_site))+
    geom_point(aes(color = AM_housing))+
    geom_line()+
    theme_minimal()
```

total detections

## Basic Info

How many identifications?

```{r}
cleaned |> filter(!is.na(probability)) |> 
    summarise(count = n())
```

Number of Species:

```{r}
unique(cleaned$species) |> length()
```

Duration of recording for each AM:

They are in only some of the deployment notes -> need to update spreadsheet

Number of ids per audiomoth and number of species

```{r}
cleaned |> group_by(deployment_num) |> 
    summarise(id_count = n(), 
    species_count = n_distinct(species)) |> 
        print(n = 30)
```

Number of ids and number of species within each housing group

```{r}
cleaned |> group_by(AM_housing) |> 
    summarise(id_count = n(), 
    species_count = n_distinct(species))
```

Mean and sd of ids and species in each housing group

```{r}
cleaned |> group_by(AM_housing, deployment_num) |> 
    summarise(id_count = n(), 
    species_count = n_distinct(species)) |> 
        ungroup() |> 
        group_by(AM_housing) |> 
        summarise(mean_id = mean(id_count),
        sd_id = sd(id_count),
        mean_sp = mean(species_count),
        sd_sp = sd(species_count))
```

## Plots:

Comparison of ids and sp for each housing group:

```{r}
dist_housing <- cleaned |> 
    group_by(AM_housing, deployment_num) |> 
    summarise(count = n(),
    num_sp = n_distinct(species))

# id plot
ggplot(dist_housing, 
aes(fct_reorder(deployment_num, count), count, fill = AM_housing))+
    geom_col()+
    coord_flip()+
    labs(y = "Number of IDs",
    x = "Deployment ID")
    theme_minimal()

# species plot
ggplot(dist_housing, 
aes(fct_reorder(deployment_num, num_sp), num_sp, fill = AM_housing))+
    geom_col()+
    coord_flip()+
    labs(y = "Number of Species",
    x = "Deployment ID")
    theme_minimal()
```

Side-by-side boxplots for ids and species:

```{r}
# IDs
ggplot(dist_housing, aes(AM_housing, count, color = AM_housing))+
    geom_boxplot()+
    geom_point(alpha = 0.5)+
    theme_minimal()
# Species
ggplot(dist_housing, aes(AM_housing, num_sp, color = AM_housing))+
    geom_boxplot()+
    geom_point(alpha = 0.5)+
    theme_minimal()
```


## Non-Paired Comparison

Use linear mixed effects models to take into account the the paired sites

Grassland: get rid of for bigger analysis

## Durability analysis

Use linear mixed effects models to take into account the paired sites. Predict the proportion of time the AM was working (working divided by length of deployment) to standardize based on deployment length

Is it worth doing a survival analysis? Saw something that it would be appriopriate for these kinds of data.

After talking to erika -- find battery life from config file and use that proportion as response.

Can we assume all the batteries were fully charged?

#### Preliminary Plot

```{r}
deploy_durability <- deploy |> mutate(last_record = mdy(last_record),
    dropoff_date = dmy(dropoff_date),
    pickup_date = dmy(pickup_date)) |> 
    group_by(deployment_num) |> 
    reframe(duration = last_record - dropoff_date,
    deployment_len = pickup_date - dropoff_date,
    durability_ratio = as.numeric(duration)/as.numeric(deployment_len),
    AM_housing = AM_housing) |> 
    ungroup() |> 
    mutate(durability_ratio = replace(durability_ratio, 
    deployment_num == "AM0057", 0)) # for the one with total failure
    
ggplot(deploy_durability, aes(AM_housing, durability_ratio, color = AM_housing))+
    geom_boxplot()+
    geom_point(alpha = 0.5)+
    theme_minimal()
```

Question: How should we think about/quantify the AM with no data on the card. 

#### Number of IDs vs. Duration of Recording

Color by baggie and case

```{r}
id_dr_plot <- cleaned |> group_by(deployment_num) |> 
    reframe(id_count = n(), 
    species_count = n_distinct(species),
    AM_housing = first(AM_housing))

ggplot(id_dr_plot, aes())
```

## accumulation curve for ids and species richness

Does the baggie group get you there quicker?

