---
title: "Preliminary Analysis"
format: html
---

## Goal:

This document looks to continue to explore/vizualize the results of the baggie/case experiment and look at some preliminary analyses.

## Set up

```{r}
rm(list = ls())
library(tidyverse)
library(here)
```

## Load in Data

```{r}
birdnet <- read_csv(here("data/updated_birdnet_results.csv"))
deploy <- read_csv(here("data/Updated_baggie_case_deployments.csv"))
```

#### Merge dataframes

```{r}
merged <- left_join(deploy, birdnet, join_by("deployment_num" == "id"))
glimpse(merged)
```

Note: there will be some rows that only have deployment info because two of the audio moths failed

We know that one of the deployments failed because of user error.

We can remove that one because it does not have to do with durability:

```{r}
cleaned <- merged |> filter(deployment_num != "AM0100") # the known AM with user error
```

#### Calculate the duration variable

```{r}
cleaned2 <- cleaned |> 
    mutate(last_record = mdy(last_record),
    dropoff_date = dmy(dropoff_date),
    pickup_date = dmy(pickup_date)) |> 
    group_by(deployment_num) |> 
    mutate(duration = last_record - dropoff_date,
    deploymet_len = pickup_date - dropoff_date) |> 
    ungroup() |> 
    mutate(duration = replace(duration, deployment_num == "AM0057", 0)) # the recorder that failed without known cause
```

## Paired data:

```{r}
paired <- cleaned2 |> filter(paired == "y") |> 
    group_by(paired_site)
```

## Basic Info

How many identifications?

```{r}
cleaned |> filter(!is.na(probability)) |> 
    summarise(count = n())
```

Number of Species:

```{r}
unique(cleaned$species) |> length()
```

Duration of recording for each AM:

They are in only some of the deployment notes -> need to update spreadsheet

Number of ids per audiomoth and number of species

```{r}
cleaned |> group_by(deployment_num) |> 
    summarise(id_count = n(), 
    species_count = n_distinct(species)) |> 
        print(n = 30)
```

Number of ids and number of species within each housing group

```{r}
cleaned |> group_by(AM_housing) |> 
    summarise(id_count = n(), 
    species_count = n_distinct(species))
```

Mean and sd of ids and species in each housing group

```{r}
cleaned |> group_by(AM_housing, deployment_num) |> 
    summarise(id_count = n(), 
    species_count = n_distinct(species)) |> 
        ungroup() |> 
        group_by(AM_housing) |> 
        summarise(mean_id = mean(id_count),
        sd_id = sd(id_count),
        mean_sp = mean(species_count),
        sd_sp = sd(species_count))
```

## Plots:

Comparison of ids and sp for each housing group:

```{r}
dist_housing <- cleaned |> 
    group_by(AM_housing, deployment_num) |> 
    summarise(count = n(),
    num_sp = n_distinct(species))

# id plot
ggplot(dist_housing, 
aes(fct_reorder(deployment_num, count), count, fill = AM_housing))+
    geom_col()+
    coord_flip()+
    labs(y = "Number of IDs",
    x = "Deployment ID")
    theme_minimal()

# species plot
ggplot(dist_housing, 
aes(fct_reorder(deployment_num, num_sp), num_sp, fill = AM_housing))+
    geom_col()+
    coord_flip()+
    labs(y = "Number of Species",
    x = "Deployment ID")
    theme_minimal()
```

Side-by-side boxplots for ids and species:

```{r}
# IDs
ggplot(dist_housing, aes(AM_housing, count, color = AM_housing))+
    geom_boxplot()+
    geom_point(alpha = 0.5)+
    theme_minimal()
# Species
ggplot(dist_housing, aes(AM_housing, num_sp, color = AM_housing))+
    geom_boxplot()+
    geom_point(alpha = 0.5)+
    theme_minimal()
```






